A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics

### 2019-03-08 pkg/colorspace/paper/Toolbox/article.Rmd
added a little more detail about some other colour packages
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@469 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

FIXME: More comparison with/discussion of other palette packages.
the viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
as provided by \proglang{R} package \pkg{viridis} [@color:viridis], and
the scientific color maps of @color:Crameri:2018 as provided by \proglang{R}
package \pkg{scico} [@color:scico]. All yield mostly well-balanced color palettes
and can be approximated reasonably well in \pkg{colorspace}
(see Section\ \ref{sec:approximations}).
A broad range of packages that provide color palettes that are "in harmony" as
they are extracted and clustered from movies, paintings, etc. However, many of these
are not so suitable for scientific visualization as they are not clearly 
qualitative/sequential/diverging. An influential package following such an
approach is \pkg{wesanderson} [@color:wesanderson]. The \pkg{paletteer}
[@color:paletteer] package aggregates these and provides them in a unified
user interface.
Aims of code in \pkg{colorspace} are general trajectories, adaptable HCL
coordinates, along with visualization and assessment of palettes (beyond
color swatches).

@@

Different kinds of limitations can be emulated using the physiologically-based model
for simulating color vision deficiency (CVD) of @color:Machado+Oliveira+Fernandes:2009:
deuteranomaly (green cone cells defective),
protanomaly (red cone cells defective), and
tritanomaly (blue cone cells defective).
While most other CVD simulations handle only dichromacy, where one of three cones is
non-functional, @color:Machado+Oliveira+Fernandes:2009 provides a unified model of
both dichromacy and anomalous trichromacy, where one cone has shifted spectral
sensitivity. As anomalous trichromacy is the most common form of color vision
deficiency, it is important to emulate along with the rarer, but more severe
dichromacy. Below we briefly describe our \proglang{R} interface to these emulation
techniques and show them in practice for a heatmap with sequential palette.
Another example with diverging palette is available at
<http://colorspace.R-Forge.R-project.org/articles/color_vision_deficiency.html>.
Finally, CVD emulation is particularly useful for bringing out why the RGB
rainbow palette is almost always a bad choice in scientific displays. See
<http://colorspace.R-Forge.R-project.org/articles/endrainbow.html> for further
illustrations.

@@

The workhorse function to emulate color vision deficiencies is `simulate_cvd()`
which can take any vector of valid \proglang{R} colors and transform them according to a
certain CVD transformation matrix and transformation equation. The
transformation matrices have been established by @color:Machado+Oliveira+Fernandes:2009 and are
provided in objects `protanomaly_cvd`, `deutanomaly_cvd`, and `tritanomaly_cvd`.
The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level
functions for simulating the corresponding kind of color blindness with a given
`severity` (calling `simulate_cvd()` internally). A severity of `1` corresponds to
dichromacy, `0` to normal color vision, and intermediate values to varying
severities of anomalous trichromacy. For further guidance on color blindness in
relation to statistical graphics see @color:Lumley:2006 which accompanies the \proglang{R}
package \pkg{dichromat} [@color:dichromat] and is based on earlier emulation techniques
[@color:Vienot+Brettel+Ott:1995; @color:Brettel+Vienot+Mollon:1997; @color:Vienot+Brettel+Mollon:1999].

@@

* \pkg{Polychrome} [@color:Polychrome]: qualitative palettes with more "categories".
* \pkg{roloc} [@color:roloc]: naming palettes.
* \pkg{colorscience} [@color:colorscience]: More color mappings.
* \pkg{paletteer} [@color:paletteer]: Aggregation of palettes.

>>>>

FIXME: More comparison with/discussion of other palette packages.
the viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
as provided by \proglang{R} package \pkg{viridis} [@color:viridis], and
the scientific color maps of @color:Crameri:2018 as provided by \proglang{R}
package \pkg{scico} [@color:scico]. All yield mostly well-balanced color palettes
and can be approximated reasonably well in \pkg{colorspace}
(see Section\ \ref{sec:approximations}).
A broad range of packages that provide color palettes that are "in harmony" as
they are extracted and clustered from movies, paintings, etc. However, many of these
are not so suitable for scientific visualization as they are not clearly 
qualitative/sequential/diverging. An influential package following such an
approach is \pkg{wesanderson} [@color:wesanderson]. The \pkg{paletteer}
[@color:paletteer] package aggregates these and provides them in a unified
user interface.
The \pkg{pals} package also provides a large number of palettes,
but some are just pre-existing palettes, and where the palettes are new
there is no description of a specific algorithm that was used to generate
them.  
Aims of code in \pkg{colorspace} are general trajectories, adaptable HCL
coordinates, along with visualization and assessment of palettes (beyond
color swatches).

@@

Using the physiologically-based model for simulating color vision deficiency (CVD)
of @color:Machado+Oliveira+Fernandes:2009 different kinds of limitations can be
emulated: deuteranope (green cone cells defective), protanope (red cone cells defective),
and tritanope (blue cone cells defective).

Below we briefly describe our \proglang{R} interface to these emulation techniques
and show them in practice for a heatmap with sequential palette. Another example
with diverging palette is available at
<http://colorspace.R-Forge.R-project.org/articles/color_vision_deficiency.html>.
Finally, CVD emulation is particularly useful for bringing out why the RGB rainbow
palette is almost always a bad choice in scientific displays. See
<http://colorspace.R-Forge.R-project.org/articles/endrainbow.html> for further
illustrations.

@@

The workhorse function to do so is `simulate_cvd()` which can take any vector
of valid \proglang{R} colors and transform them according to a certain CVD transformation
matrix and transformation equation. The transformation matrices have been established
by @color:Machado+Oliveira+Fernandes:2009 and are provided in objects `protanomaly_cvd`,
`deutanomaly_cvd`, and `tritanomaly_cvd`.
The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level functions for
simulating the corresponding kind of color blindness with a given severity (calling
`simulate_cvd()` internally).

For further guidance on color blindness in relation to statistical graphics
see @color:Lumley:2006 which accompanies the \proglang{R} package \pkg{dichromat} [@color:dichromat]
and is based on earlier emulation techniques
[@color:Vienot+Brettel+Ott:1995; @color:Brettel+Vienot+Mollon:1997; @color:Vienot+Brettel+Mollon:1999].

@@

* \pkg{Polychrome} [@color:Polychrome]: qualitative palettes with more "categories". This can generate 
  palettes with the (a large number of) 
  colours spaced far apart from each other in Luv space.
  It also has functions for visualising colour palettes in different ways 
  (https://www.biorxiv.org/content/biorxiv/early/2018/04/18/303883.full.pdf)
* \pkg{roloc} [@color:roloc]: colour conversions, but not between
  numeric colour spaces;  instead, from numeric colour spaces to
  colour names.  
  (https://stattech.wordpress.fos.auckland.ac.nz/2018/01/25/2018-01-generating-colour-names-the-roloc-package-for-r/)
* \pkg{colorscience} [@color:colorscience]: This package is for Colour
Science.  It has data sets, does more complex conversions (including
different chromatic adaptation algorithms).  
So more sophisticated than \pkg{colorspace}
in terms of flexibility and low-level detail, BUT it does not do
palette-generation.
* \pkg{paletteer} [@color:paletteer]: Aggregation of palettes.
